---
title: "Example gene set analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichmentMethods

## Intro

Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.

The dataset being used is SRP128998 and we are comparing the cells grown in normal glucose condition (control) to the high glucose condition (case).

Data are obtained from http://dee2.io/

```{r,begin}

suppressPackageStartupMessages({
library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")
})

dir.create("images")

```

## Get expression data and make an MDS plot

I'm using some RNA-seq data looking at the effect of hyperglycemia on hepatocytes.

```{r,getdata}

name="SRP128998"
mdat<-getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$trt<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
s1 <- subset(samplesheet,VPA==0)

s1 %>% kbl(caption = "sample sheet") %>% kable_paper("hover", full_width = F)

w<-getDEE2("hsapiens",samplesheet$SRR_accession,metadata=mdat,legacy = TRUE)
x<-Tx2Gene(w)
x<-x$Tx2Gene

# save the genetable for later
gt<-w$GeneInfo[,1,drop=FALSE]
gt$accession<-rownames(gt)

# counts 
x1<-x[,which(colnames(x) %in% s1$SRR_accession)]

```

Here show the number of genes in the annotation set, and those detected above the detection threshold.

```{r,filter}

# filter out lowly expressed genes
x1<-x1[which(rowSums(x1)/ncol(x1)>=(10)),]
nrow(x)
nrow(x1)

```

Now multidimensional scaling (MDS) plot to show the correlation between the datasets.
If the control and case datasets are clustered separately, then it is likely that there will be many differentially expressed genes with FDR<0.05.

```{r,mds}

plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$trt, main="MDS")

```

## Differential expression

Now run DESeq2 for control vs case.

```{r,deseq2}

y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ trt)
y <- DESeq(y)
de <- results(y)
de<-as.data.frame(de[order(de$pvalue),])
rownames(de)<-sapply(strsplit(rownames(de),"\\."),"[[",1)
head(de) %>% kbl() %>% kable_paper("hover", full_width = F)

```

Now let's have a look at some of the charts showing differential expression.
In particular, an MA plot and volcano plot.

```{r,deplots,fig.width=7,fig.height=7}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange, 
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=0.5, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=0.5, col="red")
  mtext(SUBHEADER,cex = 0.7)
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=0.5,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval", xlim=c(-6,6))
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=0.5,pch=19,col="red")
}

maplot(de,name)
make_volcano(de,name)

```

## Gene sets from Reactome

In order to perform gene set analysis, we need some gene sets. 

```{r,reactome}

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
genesets<-gmt_import("ReactomePathways.gmt")

```

## FCS with Mitch

Mitch uses rank-ANOVA statistics for enrichment detection.

```{r,mitch1}

m <- mitch_import(de,DEtype = "DEseq2", geneTable = gt)
mres <- mitch_calc(m,genesets = genesets)
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
message(paste("Number of up-regulated pathways:",length(m_up) ))
message(paste("Number of down-regulated pathways:",length(m_dn) ))
head(mres$enrichment_result,10)  %>% kbl() %>% kable_paper("hover", full_width = F)

m_up_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist > 0)[,1]
m_dn_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist < 0)[,1]

```

## ORA with clusterprofiler

Clusterprofiler uses a hypergeometric test.
Firstly I will conduct the analysis separately for up and down regulated genes and with the correct backgound (as intended by the developers).

```{r,cp1}

genesets2 <- read.gmt("ReactomePathways.gmt")

de_up <- rownames(subset(de,log2FoldChange>0,padj<0.05))
de_up <- unique(gt[which(rownames(gt) %in% de_up),1])

de_dn <- rownames(subset(de,log2FoldChange<0,padj<0.05))
de_dn <- unique(gt[which(rownames(gt) %in% de_dn),1])

de_bg <- rownames(de)
de_bg <- unique(gt[which(rownames(gt) %in% de_bg),1])

c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
c_up <- rownames(subset(c_up, p.adjust < 0.05))
       
c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

c_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none" ))
c_up_nom <- rownames(subset(c_up_nom, pvalue < 0.05))

c_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
c_dn_nom <- rownames(subset(c_dn_nom, pvalue < 0.05))

```

Now performing ORA with clusterprofiler combining up and down. (Not shown in final article)

```{r,cp2}

de_de <- rownames(subset(de,padj<0.05))
de_de <- unique(gt[which(rownames(gt) %in% de_de),1])

d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
d_de <- rownames(subset(d_de, p.adjust < 0.05))

```

Now performing ORA with clusterprofiler with whole genome background list

```{r,cp3}

de_bg <- w$GeneInfo$GeneSymbol

f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
f_up <- rownames(subset(f_up, p.adjust < 0.05))
       
f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

f_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
f_up_nom <- rownames(subset(f_up_nom, p.adjust < 0.05))

f_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
f_dn_nom <- rownames(subset(f_dn_nom, p.adjust < 0.05))

```


Now performing ORA (combining up and down gene lists) with clusterprofiler with whole genome background list

```{r,cp4}

e_de <- as.data.frame(enricher(gene = de_de, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2))
e_de <- rownames(subset(e_de, p.adjust < 0.05))

```

## Bar chart of significance

Here the idea is to classify the pathways as 1 not significant, 2 nominally significant, or 3 FDR significant

```{r,stacked1}

c_up_df <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))

n_c_up_ns <- nrow( subset(c_up_df,pvalue>0.05) )
n_c_up_nom <- nrow( subset(c_up_df,pvalue<0.05 ) )
n_c_up_fdr <- nrow( subset(c_up_df,p.adjust<0.05) )

c_dn_df <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))

n_c_dn_ns <- nrow( subset(c_dn_df,pvalue>0.05) )
n_c_dn_nom <- nrow( subset(c_dn_df,pvalue<0.05 ) )
n_c_dn_fdr <- nrow( subset(c_dn_df,p.adjust<0.05) )

n_m_up <- length(m_up)
n_m_dn <- length(m_dn)

n_m_up_nom <- length(m_up_nom)
n_m_dn_nom <- length(m_dn_nom)

par(mar=c(5,10,5,2))

ngenes <- c("ORA up fdr"=n_c_up_fdr,"ORA up nom"=n_c_up_nom,
  "ORA dn fdr"=n_c_dn_fdr,"ORA dn nom"=n_c_dn_nom,
  "FCS up fdr"=n_m_up,"FCS up nom"=n_m_up_nom,
  "FCS dn fdr"=n_m_dn,"FCS dn nom"=n_m_dn_nom )

barplot(ngenes, horiz=TRUE,las=1)

n_c_fdr <- length(union(c_dn,c_up))
n_c_nom <- length(union(c_dn_nom,c_up_nom))
n_m_fdr <- length(subset(mres$enrichment_result,p.adjustANOVA<0.05 )[,1])
n_m_nom <- length(subset(mres$enrichment_result,pANOVA<0.05 )[,1])

par(mar=c(5,5,5,2))

ngenes <- c("ORA FDR<0.05"=n_c_fdr,"ORA p<0.05"=n_c_nom,"FCS FDR<0.05"=n_m_fdr,"FCS p<0.05"=n_m_nom)
barplot(ngenes,ylab="no. gene sets")
text((0:3*1.2)+0.7,ngenes-50,labels=ngenes,cex=1.2)

pdf("images/fdr1.pdf",width=4,height=4)
par(mar=c(5,7,2,2))
barplot(ngenes,xlab="no. gene sets",horiz=TRUE,las=1)
text( ngenes - 70, (0:3*1.2)+0.7,labels=ngenes,cex=1)
dev.off()

```

## Venn diagram comparison

The Venn (or Euler to be more correct) diagram is useful to visualise the overlaps between sets.

```{r,venn1}

par(cex.main=0.5)

par(mar=c(2,2,2,2))

v0 <- list("ORA up"=c_up,"ORA dn"=c_dn,
           "ORA comb" = d_de)

plot(euler(v0),quantities = TRUE, edges = "gray", main="effect of combining up and down regulated genes")

v1 <- list("FCS up"=m_up, "FCS dn"=m_dn,
           "ORA up"=c_up,"ORA dn"=c_dn)
  
plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS compared to ORA")

v2 <- list("ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn )

plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")

vx <- list("ORA up"=c_up,"ORA dn"=c_dn,
           "ORA comb" = d_de, "ORA* comb" = e_de)

plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")

v3 <- list("ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn ,
           "FCS up"=m_up, "FCS dn"=m_dn)

png("images/fcs_ora1.png")
plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
dev.off()

png("images/orabg1.png")
plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
dev.off()

png("images/oracomb1.png")
plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")
dev.off()

pdf("images/fcs_ora1.pdf",width=4,height=4)
plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
dev.off()

pdf("images/orabg1.pdf",width=4,height=4)
plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
dev.off()

pdf("images/oracomb1.pdf",width=4,height=4)
plot(euler(vx),quantities = TRUE, edges = "gray", main="combining up and down genes and whole genome bg*")
dev.off()

```

In the next Venn I will omit combined analysis and include nominal p-value sets.

```{r,venn1}

par(cex.main=0.5)

par(mar=c(2,2,2,2))

v1 <- list("FCS up"=m_up, "FCS dn"=m_dn,
           "ORA up"=c_up,"ORA dn"=c_dn)

plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS compared to ORA")

v0 <- list("FDR up"=m_up, "FDR dn"=m_dn,
           "Nom up"=m_up_nom,"Nom dn"=m_dn_nom)

plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on FCS results")

v0 <- list("FDR up"=c_up, "FDR dn"=c_dn,
           "Nom up"=c_up_nom,"Nom dn"=c_dn_nom)

plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on ORA results")

ora_nom <- union(c_up_nom,c_dn_nom)
ora_fdr <- union(c_up,c_dn)
fcs_nom <- union(m_up_nom,m_dn_nom)
fcs_fdr <- union(m_up,m_dn)

v3 <- list("ORA nom"=ora_nom, "ORA FDR"=ora_fdr,
           "FCS nom"=fcs_nom,"FCS FDR"=fcs_fdr)

plot(euler(v3),quantities = TRUE, edges = "gray", main="Effect of FDR correction")




v2 <- list("ORA up"=c_up,"ORA dn"=c_dn,
           "ORA* up"=f_up,"ORA* dn"=f_dn )

plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")



png("images/fcs_ora1.png")
plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
dev.off()

png("images/orabg1.png")
plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
dev.off()

pdf("images/fcs_ora1.pdf",width=4,height=4)
plot(euler(v1),quantities = TRUE, edges = "gray", main="FCS vs ORA")
dev.off()

pdf("images/orabg1.pdf",width=4,height=4)
plot(euler(v2),quantities = TRUE, edges = "gray", main="Effect of inappropriate background* (whole genome)")
dev.off()

```

## Jaccard calculation

```{r,jacc1}

# ORA vs ORA combined
dc <- length(intersect(d_de, c(c_up,c_dn))) / length(union(d_de, c(c_up,c_dn)))

# ORA vs ORA* combined
ec <- length(intersect(e_de, c(c_up,c_dn))) / length(union(e_de, c(c_up,c_dn)))

# FCS vs ORA
cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))

m_up <- gsub("^","up ",m_up)
m_dn <- gsub("^","dn ",m_dn)
m_de <- union(m_up,m_dn)

c_up <- gsub("^","up ",c_up)
c_dn <- gsub("^","dn ",c_dn)
c_de <- union(c_up,c_dn)

f_up <- gsub("^","up ",f_up)
f_dn <- gsub("^","dn ",f_dn)
f_de <- union(f_up,f_dn)

# ORA vs ORA*
cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))

# FCS vs ORA*
mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))

dat <- c("FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf, "ORA vs ORA comb"=dc, "ORA vs ORA* comb"=ec)

dat
saveRDS(dat,file = "ex1dat.rds")

par(mar=c(5,10,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))

pdf("images/jacbar.pdf",width=4,height=3)
par(mar=c(5,9,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.8) , main=name)
text( x=rev(dat)+0.08 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()

png("images/jacbar.png")
par(mar=c(5,10,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()

```

As per reviewer suggestion, remove the comparison of combined and separated ORA.

I will include using nominal p-value.

```{r,jacc2}

# FCS vs ORA
cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))

m_up <- gsub("^","up ",m_up)
m_dn <- gsub("^","dn ",m_dn)
m_de <- union(m_up,m_dn)

c_up <- gsub("^","up ",c_up)
c_dn <- gsub("^","dn ",c_dn)
c_de <- union(c_up,c_dn)

f_up <- gsub("^","up ",f_up)
f_dn <- gsub("^","dn ",f_dn)
f_de <- union(f_up,f_dn)

# ORA vs ORA*
cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))

# FCS vs ORA*
mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))

dat <- c("FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)

dat
saveRDS(dat,file = "ex1dat.rds")

par(mar=c(5,10,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))

pdf("images/jacbar2.pdf",width=4,height=3)
par(mar=c(5,9,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.8) , main=name)
text( x=rev(dat)+0.08 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()

png("images/jacbar2.png")
par(mar=c(5,10,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()

```

## Session information

```{r,session}

sessionInfo()

```
